---
title: "Demultiplex report"
# author: "Tram"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
   rmd: "report_demultiplex.Rmd"
output: html_document
---

<style>
.main-container {
    width: 70%;
    max-width: unset;
}
</style>

The report provides plots in order to compare demultiplexing results from different runs, and from different demultiplexers (guppy and deepbinner), and for each genome (if the table barcode by genome is provided).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
if (!requireNamespace("DT", quietly = TRUE)) install.packages('DT', repos = "https://cloud.r-project.org")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages('ggplot2', repos = "https://cloud.r-project.org")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages('dplyr', repos = "https://cloud.r-project.org")
if (!requireNamespace("stringr", quietly = TRUE)) install.packages('stringr', repos = "https://cloud.r-project.org")
if (!requireNamespace("S4Vectors", quietly = TRUE)) install.packages('S4Vectors', repos = "https://cloud.r-project.org")
# if (!requireNamespace("parallel", quietly = TRUE)) install.packages('parallel', repos = "https://cloud.r-project.org")
if (!requireNamespace("Biostrings", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager", repos = "https://cloud.r-project.org")

  BiocManager::install("Biostrings", update = F, ask = F)
}

library(stringr)
library(ggplot2)
library(dplyr)
library(DT)
library(Biostrings)
library(S4Vectors)
library(parallel)
```


```{r demux}
fastq_folder <- snakemake@params$fastq # "/data3/projects/xanthopore/baotram/test/results_guppy/reads_per_genome/fastq"
demux_fastq <- list.files(fastq_folder, ".*\\.fastq.gz", full.names = T)
demux_fastq <- demux_fastq[grep("porechop|filtlong", demux_fastq, invert = T)]

demux <- do.call(rbind, mclapply(demux_fastq, function(fg) {
  fastq <- readDNAStringSet(fg, format = "fastq", seek.first.rec = T, with.qualities = T)
  # phred score
  phred <- PhredQuality(mcols(fastq)$qualities)
  scores <- unlist(lapply(as(phred, "IntegerList"), mean))

  dt <- data.frame("read_id" = gsub(" .+", "", names(fastq)),
                   "run_id" = gsub("(.+ runid=| sampleid.+)", "", names(fastq)),
                   "read_length" = width(fastq),
                   "phred.score" = scores,
                   "Genome_ID" = gsub("\\.fastq.gz", "", basename(fg)),
                   "Step" = "demultiplex",
                   "id" = names(fastq),
                   stringsAsFactors = F)
  return(dt)
}))
```

```{r sequencing_summary}
demux_folder <- snakemake@params$demultiplex # "/data3/projects/xanthopore/baotram/test/results_guppy/demultiplex"
summaries <- list.files(demux_folder, "sequencing_summary.txt", recursive = T, full.names = T)
sum <- do.call(rbind, lapply(summaries, function(s) {
  sum <- read.table(s, header = T)
  splitted_path <- unlist(strsplit(dirname(s), "/", fixed = T))
  sum <- data.frame(sum,
                    "ONT_Barcode" = splitted_path[length(splitted_path)],
                    "Run_ID" = splitted_path[length(splitted_path) - 1],
                    "Demultiplexer" = splitted_path[length(splitted_path) - 2],
                    stringsAsFactors = F)
  return(sum)
}))
```

```{r demux_summary}
demux_sum <- merge(demux, sum[c("read_id", "run_id", "ONT_Barcode", "Run_ID", "Demultiplexer")], by = c("read_id", "run_id"))
```

```{r postdemux}
postdemux_folder <- snakemake@params$postdemux # "/data3/projects/xanthopore/baotram/test/results_guppy/reads_per_genome/fastq_porechop_filtlong"
if (any(postdemux_folder != demux_folder)) {
  postdemux_fastq <- list.files(postdemux_folder, ".*\\.fastq.gz", full.names = T)
  postdemux_fastq <- postdemux_fastq[grep("porechop|filtlong", postdemux_fastq)]

  postdemux <- do.call(rbind, mclapply(postdemux_fastq, function(fg) {
    fastq <- readDNAStringSet(fg, format = "fastq", seek.first.rec = T, with.qualities = T)
    # phred score
    phred <- PhredQuality(mcols(fastq)$qualities)
    scores <- unlist(lapply(as(phred, "IntegerList"), mean))

    dt <- data.frame("read_id" = gsub(" .+", "", names(fastq)),
                     "run_id" = gsub("(.+ runid=| sampleid.+)", "", names(fastq)),
                     "read_length" = width(fastq),
                     "phred.score" = scores,
                     # "Genome_ID" = gsub("\\.fastq.gz", "", basename(fg)),
                     "Step" = gsub("fastq", "demultiplex", basename(dirname(fg))),
                     "id" = names(fastq),
                     stringsAsFactors = F)
    return(dt)
  }))

  postdemux <- merge(postdemux, demux[c("read_id", "run_id", "Genome_ID")], by = c("read_id", "run_id"))

  postdemux_sum <- merge(postdemux, sum[c("read_id", "run_id", "ONT_Barcode", "Run_ID", "Demultiplexer")], by = c("read_id", "run_id"))
}
```

```{r all_qc}
if (any(postdemux_folder == demux_folder)) {
  all_sum <- demux_sum %>% dplyr::mutate(Step = NA)
} else {
  all_sum <- rbind(demux_sum, postdemux_sum)
}


all_sum$Demultiplexer <- factor(all_sum$Demultiplexer, levels = c("guppy", "deepbinner"), ordered = F)
# all_sum$Step <- factor(all_sum$Step, levels = c(gsub("fastq", "demultiplex", basename(postdemux_folder)), "demultiplex"), ordered = F)

find_n50 <- function(x) {
  thredshold <- sum(x)/2
  test <- 0
  for (l in sort(x)) {
    test <- test + l
    if (test > thredshold) break
  }
  return(l)
}

byrun <- all_sum %>%
  # dplyr::filter(Step == "demultiplex") %>%
  dplyr::group_by(Step, Demultiplexer, Run_ID, ONT_Barcode) %>%
  dplyr::summarise(total.gigabases = sum(read_length)/10**9,
                   total.reads = length(read_length),
                   mean.length = mean(read_length),
                   median.length = median(read_length),
                   N50.length = find_n50(read_length),
                   mean.quality.score = mean(phred.score),
                   median.quality.score = median(phred.score),
                   N50.quality.score = find_n50(phred.score))
```


# {.tabset .tabset-fade }
## by run {.tabset .tabset-fade }

```{r}
if (nlevels(factor(byrun$Step)) > 1) {
  cat("FILTERING PARAMS\n\n")
  if ("porechop" %in% names(snakemake@config)) print(snakemake@config["porechop"])
  filtlongs <- names(snakemake@config)[grep("filtlong.+", names(snakemake@config))]
  if (length(filtlongs) > 0) print(snakemake@config[filtlongs])
}
```

### Total gigabases

```{r run.total.gigabases, fig.width = 16, fig.height = 8}
ggplot(byrun, aes(x = ONT_Barcode, y = total.gigabases, fill = Demultiplexer, alpha = Step)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0), stat = "identity") +
  scale_fill_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  facet_wrap(vars(Run_ID), scales = "fixed", ncol = 3, strip.position = "right") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.direction = "horizontal", legend.position = "top")
```

### Total reads

```{r run.total.reads, fig.width = 16, fig.height = 8}
ggplot(byrun, aes(x = ONT_Barcode, y = total.reads, fill = Demultiplexer, alpha = Step)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
  scale_fill_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  facet_wrap(vars(Run_ID), scales = "fixed", ncol=3, strip.position = "right") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.direction = "horizontal", legend.position = "top")
```

### Mean length

```{r run.mean.length, fig.width = 16, fig.height = 8}
ggplot(byrun, aes(x = ONT_Barcode, y = mean.length, fill = Demultiplexer, alpha = Step)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
  scale_fill_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  facet_wrap(vars(Run_ID), scales = "fixed", ncol = 3, strip.position = "right") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.direction = "horizontal", legend.position = "top")
```

### Median length

```{r run.median.length, fig.width = 16, fig.height = 8}
ggplot(byrun, aes(x = ONT_Barcode, y = median.length, fill = Demultiplexer, alpha = Step)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
  scale_fill_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  facet_wrap(vars(Run_ID), scales = "fixed", ncol = 3, strip.position = "right") +
  theme_light() +
  theme(axis.text.x=element_text(angle=90,hjust=1),
        legend.direction = "horizontal", legend.position = "top")
```

### N50 length

```{r run.N50.length, fig.width = 16, fig.height = 8}
ggplot(byrun, aes(x = ONT_Barcode, y = N50.length, fill = Demultiplexer, alpha = Step)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
  scale_fill_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  facet_wrap(vars(Run_ID), scales = "fixed", ncol = 3, strip.position = "right") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.direction = "horizontal", legend.position = "top")
```

### Mean quality score

```{r run.mean.quality, fig.width = 16, fig.height = 8}
ggplot(byrun, aes(x = ONT_Barcode, y = mean.quality.score, fill = Demultiplexer, alpha = Step)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
  scale_fill_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  facet_wrap(vars(Run_ID), scales = "fixed", ncol = 3, strip.position = "right") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.direction = "horizontal", legend.position = "top")
```

### Median quality score

```{r run.median.quality, fig.width = 16, fig.height = 8}
ggplot(byrun, aes(x = ONT_Barcode, y = median.quality.score, fill = Demultiplexer, alpha = Step)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
  scale_fill_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  facet_wrap(vars(Run_ID), scales = "fixed", ncol = 3, strip.position = "right") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.direction = "horizontal", legend.position = "top")
```

### N50 quality score

```{r run.n50.quality, fig.width = 16, fig.height = 8}
ggplot(byrun, aes(x = ONT_Barcode, y = N50.quality.score, fill = Demultiplexer, alpha = Step)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
  scale_fill_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  facet_wrap(vars(Run_ID), scales = "fixed", ncol = 3, strip.position = "right") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.direction = "horizontal", legend.position = "top")
```


### Quality score distribution

```{r run.quality.distribution, fig.width = 16, fig.height = 8}
ggplot(all_sum, aes(x = round(phred.score))) +
  geom_line(aes(alpha = Step, color = Demultiplexer), stat = "count") +
  facet_grid(cols = vars(Run_ID), rows = vars(Demultiplexer), scales = "fixed") +
  scale_color_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  theme_light() +
  theme(axis.text.x=element_text(angle=90,hjust=1),
        legend.direction = "horizontal", legend.position = "top")
```

## by genome {.tabset .tabset-fade }

```{r}
bcgenome_tsv = snakemake@params$barcode_by_genome # "/data3/projects/xanthopore/baotram/baseDmux/barcodeByGenome_sample.tsv"
if (file.exists(bcgenome_tsv)) {
  bcgenome <- read.table(bcgenome_tsv, header = T, sep = "\t", stringsAsFactors = F)
  bygenome <- merge(byrun, bcgenome, by = c("Demultiplexer", "Run_ID", "ONT_Barcode"))
}
```

```{r}
if (file.exists(bcgenome_tsv)) {
  if (nlevels(factor(bygenome$Step)) > 1) {
  cat("FILTERING PARAMS\n\n")
  if ("porechop" %in% names(snakemake@config)) print(snakemake@config["porechop"])
  filtlongs <- names(snakemake@config)[grep("filtlong.+", names(snakemake@config))]
  if (length(filtlongs) > 0) print(snakemake@config[filtlongs])
  }
}
```

### Total gigabases

```{r genome.total.gigabases, fig.width = 16, fig.height = 10}
if (file.exists(bcgenome_tsv)) {
  p <- ggplot(bygenome, aes(x = Genome_ID, y = total.gigabases, fill = Run_ID)) +
    geom_col(position = position_stack(), color = "gray") +
    geom_text(aes(label = ONT_Barcode), color = "#b1bed5", position = position_stack(), vjust = 1.2, size = 3.88*16/nlevels(as.factor(bygenome$Genome_ID))) +
    scale_fill_viridis_d(option = "C") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.direction = "vertical", legend.position = "top")

if (!all(is.na(bygenome$Step))) p <- p + facet_wrap(vars(Step), nrow = nlevels(factor(bygenome$Step)))
  p
}
```

### Total reads

```{r genome.total.reads, fig.width = 16, fig.height = 10}
if (file.exists(bcgenome_tsv)) {
  p <- ggplot(bygenome, aes(x = Genome_ID, y = total.reads, fill=Run_ID)) +
    geom_col(position = position_stack(), color = "gray") +
    geom_text(aes(label = ONT_Barcode), color = "#b1bed5", position = position_stack(), vjust = 1.2, size = 3.88*16/nlevels(as.factor(bygenome$Genome_ID))) +
    scale_fill_viridis_d(option = "C") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.direction = "vertical", legend.position = "top")

  if (!all(is.na(bygenome$Step))) p <- p + facet_wrap(vars(Step), nrow = nlevels(factor(bygenome$Step)))
  p
}
```

### Mean length

```{r genome.mean.length, fig.width = 16, fig.height = 10}
if (file.exists(bcgenome_tsv)) {
  p <- ggplot(bygenome, aes(x = Genome_ID, y = mean.length, fill = Run_ID)) +
    geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
    geom_text(aes(label = ONT_Barcode), color = "#b1bed5", position = position_dodge2(width = 1, preserve = "single", padding = 0), angle = 90, vjust= .5, hjust = 1, size = 3.88*8/nlevels(as.factor(bygenome$Step))) +
    scale_fill_viridis_d(option = "C") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.direction = "vertical", legend.position = "top")

  if (!all(is.na(bygenome$Step))) p <- p + facet_wrap(vars(Step), nrow = nlevels(factor(bygenome$Step)))
  p
}
```

### Median length

```{r genome.median.length, fig.width = 16, fig.height = 10}
if (file.exists(bcgenome_tsv)) {
  p <- ggplot(bygenome, aes(x = Genome_ID, y = median.length, fill = Run_ID)) +
    geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
    geom_text(aes(label = ONT_Barcode), color = "#b1bed5", position = position_dodge2(width = 1, preserve = "single", padding = 0), angle = 90, vjust= .5, hjust = 1, size = 3.88*8/nlevels(as.factor(bygenome$Step))) +
    scale_fill_viridis_d(option = "C") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.direction = "vertical", legend.position = "top")

  if (!all(is.na(bygenome$Step))) p <- p + facet_wrap(vars(Step), nrow = nlevels(factor(bygenome$Step)))
  p
}
```

### N50 length

```{r genome.N50.length, fig.width = 16, fig.height = 10}
if (file.exists(bcgenome_tsv)) {
  p <- ggplot(bygenome, aes(x = Genome_ID, y = N50.length, fill = Run_ID)) +
    geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
    geom_text(aes(label = ONT_Barcode), color = "#b1bed5", position = position_dodge2(width = 1, preserve = "single", padding = 0), angle = 90, vjust= .5, hjust = 1, size = 3.88*8/nlevels(as.factor(bygenome$Step))) +
    scale_fill_viridis_d(option = "C") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.direction = "vertical", legend.position = "top")

  if (!all(is.na(bygenome$Step))) p <- p + facet_wrap(vars(Step), nrow = nlevels(factor(bygenome$Step)))
  p
}
```

### Mean quality score

```{r genome.mean.quality, fig.width = 16, fig.height = 10}
if (file.exists(bcgenome_tsv)) {
  p <- ggplot(bygenome, aes(x = Genome_ID, y = mean.quality.score, fill = Run_ID)) +
    geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
    geom_text(aes(label = ONT_Barcode), color = "#b1bed5", position = position_dodge2(width = 1, preserve = "single", padding = 0), angle = 90, vjust= .5, hjust = 1, size = 3.88*8/nlevels(as.factor(bygenome$Step))) +
    scale_fill_viridis_d(option = "C") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.direction = "vertical", legend.position = "top")

  if (!all(is.na(bygenome$Step))) p <- p + facet_wrap(vars(Step), nrow = nlevels(factor(bygenome$Step)))
  p
}
```

### Median quality score

```{r genome.median.quality, fig.width = 16, fig.height = 10}
if (file.exists(bcgenome_tsv)) {
  p <- ggplot(bygenome, aes(x = Genome_ID, y = median.quality.score, fill = Run_ID)) +
    geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
    geom_text(aes(label = ONT_Barcode), color = "#b1bed5", position = position_dodge2(width = 1, preserve = "single", padding = 0), angle = 90, vjust= .5, hjust = 1, size = 3.88*8/nlevels(as.factor(bygenome$Step))) +
    scale_fill_viridis_d(option = "C") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.direction = "vertical", legend.position = "top")

  if (!all(is.na(bygenome$Step))) p <- p + facet_wrap(vars(Step), nrow = nlevels(factor(bygenome$Step)))
  p
}
```

### N50 quality score

```{r genome.n50.quality, fig.width = 16, fig.height = 10}
if (file.exists(bcgenome_tsv)) {
  p <- ggplot(bygenome, aes(x = Genome_ID, y = N50.quality.score, fill = Run_ID)) +
    geom_col(position = position_dodge2(preserve = "single", padding = 0)) +
    geom_text(aes(label = ONT_Barcode), color = "#b1bed5", position = position_dodge2(width = 1, preserve = "single", padding = 0), angle = 90, vjust= .5, hjust = 1, size = 3.88*8/nlevels(as.factor(bygenome$Step))) +
    scale_fill_viridis_d(option = "C") +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.direction = "vertical", legend.position = "top")

  if (!all(is.na(bygenome$Step))) p <- p + facet_wrap(vars(Step), nrow = nlevels(factor(bygenome$Step)))
  p
}
```

### Quality score distribution

```{r genome.quality.distribution, fig.width = 16, fig.height = 10}
ggplot(all_sum, aes(x = round(phred.score))) +
  geom_line(aes(alpha = Step, color = Demultiplexer), stat = "count") +
  facet_grid(cols = vars(Genome_ID), rows = vars(Demultiplexer), scales = "fixed") +
  scale_color_viridis_d(option = "E", end = .85) +
  scale_alpha_discrete(range = c(.3, 1)) +
  theme_light() +
  theme(axis.text.x=element_text(angle=90,hjust=1),
        legend.direction = "horizontal", legend.position = "top")
```

## data

```{r barcode_table}
if (file.exists(bcgenome_tsv)) {
  datatable(bygenome)
} else {
  datatable(byrun)
}
```

```{r save_workspace, include=FALSE}
if (file.exists(bcgenome_tsv)) {
  write.table(bygenome, file.path(snakemake@params$outpath, "demultiplex_report.tsv"), sep = "\t", row.names = F, quote = F)
} else {
  write.table(byrun, file.path(snakemake@params$outpath, "demultiplex_report.tsv"), sep = "\t", row.names = F, quote = F)
}
image_path <- file.path(snakemake@params$outpath, "demultiplex_report.RData")
remove(p)
save.image(file = image_path, compress = TRUE)
```
